Multi-dimensionality of tree communities structure host-parasitoid networks and their phylogenetic composition

Version 1

Abstract

Abstract

Environmental factors can influence ecological networks, but these effects are poorly understood in the realm of the phylogeny of host-parasitoid interactions. Especially, we lack a comprehensive understanding of the ways that biotic factors, including plant diversity, tree identity, genetic diversity, overall community composition of higher trophic levels, and abiotic factors such as microclimate, interact to determine host–parasitoid network structure and host–parasitoid community dynamics. To address this, we leveraged a five-year dataset of trap-nesting bees and wasps and their parasitoids collected in a highly-controlled, large-scale subtropical tree biodiversity experiment. We tested for effects of tree species richness, tree phylogenetic and functional diversity, and taxonomic and phylogenetic composition on taxonomic, phylogenetic, and network composition of both host and parasitoid communities. We show that multiple components of tree diversity, tree composition, and canopy cover impacted both, taxonomic and phylogenetic composition of hosts and parasitoids. Generally, top-down control was stronger than bottom-up control via phylogenetic association between hosts and parasitoids, reflecting non-randomly structured interactions between phylogenetic trees of hosts and parasitoids. Further, host-parasitoid network structure was influenced by tree species richness, tree phylogenetic diversity, and canopy cover. Our study indicates that the composition of higher trophic levels and corresponding interaction networks are determined by habitat structure and heterogeneity, which is maintained by trees and especially via phylogenetic links in species-rich ecosystems.

Introduction

Understanding the ecological consequences of biodiversity loss is an increasingly important task in ecology, given the ongoing biodiversity crisis (Isbell et al. 2022). Representing the interdependencies among organisms, ecological networks reflect whether and how species interact with each other across trophic levels, playing an indispensable role in assessing ecosystem stability and integrity (De Ruiter et al. 1995, Harvey et al. 2017). Changes in network structure usually coincide with variations in diversity and community of trophic levels, which could be in turn affected by the changes in producers via trophic cascades (Barnes et al. 2018, Gonzalez et al. 2020). However, we still lack a generalizable framework for how these networks and especially their phylogenetic interdependencies respond to basal changes of biodiversity loss in ecosystems, such as changes in the tree diversity of forests (Tylianakis et al. 2008, Grossman et al. 2018). To better understand species codependence and its role for biodiversity conservation, we must further study the complex dynamics of networks from multiple angles (Tittensor et al. 2014, Brondizio et al. 2019). Interactions can be viewed from a top-down or bottom-up perspective. Higher trophic levels can impact lower trophic levels through antagonistic species interactions, while lower trophic levels can provide resources to support higher trophic levels, Previous studies have shown there might be asymmetric effects between top-down and bottom-up control by different mediators (Vidal and Murphy 2018), with both factors shaping multitrophic communities together(Hunter et al. 1992). Moreover, the diversity and community of higher trophic levels could be driven by environmental microclimate, which could be determined by plant structuring, such as canopy cover (Fornoff et al. 2021, Perlík et al. 2023). Understanding how these forces respond to changing biodiversity is therefore imperative to predict how ongoing environmental changes will impact the functioning and stability of ecosystems (Hines et al. 2019).

A crucial network that unites bottom-up and top-down processes in many ecosystems and is prone to strong alterations due to environmental change is the interaction network between parasitoids and their hosts (Tylianakis et al. 2006, Jeffs and Lewis 2013). Insect parasitoids attack and feed on and eventually kill their insect hosts (Godfray and Godfray 1994). Parasitoids are thought to be particularly sensitive to environmental changes, because species in higher trophic levels usually have smaller population sizes and the potential specialization of their hosts may cascade up to impact the parasitoids’ presence or absence. Therefore, insect host-parasitoid systems are ideal for studying the relationships between habitat-level changes and species interactions (Jeffs and Lewis 2013). Previous studies on host-parasitoid interactions mainly focused on the influence of abiotic factors, such as elevation and habitat structure (e.g. Valladares et al. 2012, Maunsell et al. 2015, Grass et al. 2018), or sometimes interaction structure (e.g. Cagnolo et al. 2011). However, the role of multiple components of plant diversity in modifying host-parasitoid interaction networks, as key biotic determinants of overall ecosystem structure, remains relatively little explored. Recent studies mainly focused on basic diversity associations (Ebeling et al. 2012, Schuldt et al. 2019, Guo et al. 2021). These studies have demonstrated both direct and indirect effects of plant diversity on both hosts and parasitoids, possibly via increased niche space and resource availability (Guo et al. 2021). Nevertheless, how these patterns propagate to their interaction networks is still unclear. Moreover, the effects of changing plant diversity are not always obvious when looking at plant species richness, and other diversity components such as plant phylogenetic diversity have been shown to better predict diversity-dependent bottom-up effects on host-parasitoid networks (e.g. Staab et al. 2016). It is especially important to take phylogenetic dependences into account when it comes to the phylogenetic structure within and across trophic levels (Webb et al. 2002, Emerson and Gillespie 2008). This makes it vital to account for multiple dimensions of biodiversity and relevant trophic interactions (e.g. taxonomic, phylogenetic, functional; Peralta et al. 2015, Volf et al. 2017, Wang et al. 2020). These components might jointly affect host-parasitoid networks in a system with high species diversity. Forests have garnered special attention lately, because they represent complex and large ecosystems susceptible to global change (add citation here). Understanding how such relationships modulate the effects of tree diversity loss on the structure and interaction strength of host-parasitoid networks clearly requires further study (Staab et al. 2016, Fornoff et al. 2019).

Here, we leverage standardized trap nests for solitary cavity-nesting bees, wasps, and their parasitoids in a large-scale subtropical forest biodiversity experiment to test how multiple dimensions of tree diversity and community composition influence host-parasitoid network structure. A multi-faceted approach is particularly important when considering that associations between trophic levels might be non-random and phylogenetically structured (Volf et al. 2018, Wang et al. 2020). We aimed to quantify the strength and stability of associations between hosts and parasitoids to both discern the primary components of the diversity and composition of tree communities that affect higher trophic level interactions We expected that (a) multiple tree community metrics, such as taxonomic, phylogenetic and functional diversity and phylogenetic and species community composition can structure host and parasitoid community compositions, especially via phylogenetic processes, as species interactions often show phylogenetic conservatism (e.g. Pellissier et al. 2013, Peralta et al. 2015). Further, we hypothesized that (b) host-parasitoid networks will be more complex and stable with increasing tree species richness due to potential links from higher richness of hosts and parasitoids promoted by more tree species, and (c) both, community and interaction network changes can also be related to abiotic factors, such as canopy cover, which might play a role in structuring hymenopteran communities (Haddad et al. 2011, Fornoff et al. 2021). By better understanding these dynamics, we can begin building a generalized framework for understanding host-parasitoid interactions in forest ecosystems.

Results

Overall, 34,398 brood cells were collected from 13,267 tubes across five years of sampling (2015, 2016, 2018, 2019, and 2020). Six families of hosts and seventeen families of parasitoids were identified. Among them, we found 56 host species (12 bees and 44 wasps) and 50 parasitoid species (38 Hymenoptera and 12 Diptera). The full species list and their abundances are given in Table S1.

Community composition of hosts and parasitoids

Host species composition was significantly related to the species and phylogenetic composition of the tree and parasitoid communities (NMDS scores), as well as to canopy cover, tree MPD, elevation, and eastness (Fig. 1a, Table 1, Table S2). Parasitoid species composition was significantly associated with host phylogenetic diversity, tree functional diversity, tree MPD, eastness, and elevation, and was significantly related to tree species composition, host species composition, and canopy cover (Fig. 1b, Table 1, Table S3). Host phylogenetic composition was affected by tree species composition, tree MPD, tree functional diversity, canopy cover, eastness, elevation and was especially obviously affected by parasitoid species and phylogenetic composition (Fig. 1c, Table 1, Table S4). For parasitoid phylogenetic composition, significant relationships were found with tree species and phylogenetic composition, host species composition, tree functional diversity, canopy cover, elevation, and eastness (Fig. 1d, Table 1, Table S5).

Ordination plot of the non-metric multidimensional scaling (NMDS) analysis of (a) host taxonomic composition, (b) parasitoid taxonomic composition, (c) host phylogenetic composition, and (d) parasitoid phylogenetic composition across the study plots (filled circles) in the BEF-China experiment. Stress = 0.23, 0.23, 0.24 and 0.20, respectively. Arrows indicate significant (at p < 0.05) correlations of environmental variables with NMDS axis scores. Lengths of arrows are proportional to the strength of the correlations. Red crosses refer to the host or parasitoid species in each community. See Table S2-S5 in the Supplementary Material for abbreviations and statistical values.

An effect of host composition on the composition of the parasitoid communities was further indicated by a significant parafit test (p = 0.032), suggesting nonrandom associations in the phylogenetic structure of parasitoid and host communities (Fig. 2, Fig. S3).

Dendrogram of phylogenetic congruence for the host species (below) and associated parasitoid species (above) recorded in the study. Each rectangle represents a different superfamily (for host species) or family (for parasitoid species). H1: Pompilidae H2: Apoidea H3: Vespidae; P1: Sarcophagidae P2: Phoridae P3: Bombyliidae P4: Trigonalyidae P5: Mutillidae P6: Megachilidae P7: Chrysididae P8: Ichneumonidae P9: Chalcidoidea. The trophic network of hosts and parasitoids was non-randomly structured (parafit test: P = 0.032). Host and parasitoid species names are given in Fig. S3.

Host-parasitoid network associations

The linear regression model results showed that host vulnerability, linkage density, and robustness of parasitoids were significantly negatively related to tree species richness, while remaining unaffected by other environmental covariables (Table 2, Fig. 3; except for elevation, which was marginally significantly related to robustness). Interaction evenness was significantly negatively associated with canopy cover, and interaction evenness was also negatively related to eastness (Fig. 4c; Table 2).. Generality was only marginally associated with canopy cover, and was not related to tree species richness or the other environmental factors. In the alternative models (tree species richness replaced by tree MPD), vulnerability and linkage density were significantly positively related to tree MPD (Fig. 4a, 4b; Table S6), while parasitoid robustness was negatively related to tree MPD (Fig. S4, Table S6). The results of other network metrics (generality and interaction evenness) were consistent with those of the primary models The results of the null model analysis suggested that our metrics calculated by the observed network were significantly different from a random distribution (72, 71, and 77 out of 85 values for generality, vulnerability, and linkage density, respectively; all values for robustness, and interaction evenness), strongly demonstrating that interactions between species were not driven by random processes.

Community-level relationships of network between tree species richness and (a) vulnerability, (b) linkage density, and (c) robustness of parasitoids. Values were adjusted for covariates of the final regression model. Regression lines (with 95% confidence bands) show significant (p < 0.05) relationships. Note that axes are on a log scale for tree species richness.

Community-level relationships of network between tree mean pairwise phylogenetic distance and (a) vulnerability and (b) linkage density and community-level relationships of network between canopy cover and (c) interaction evenness. Values were adjusted for covariates of the final regression model. Regression lines (with 95% confidence bands) show significant (p < 0.05) relationships.

Discussion

Our study demonstrates that tree species richness and phylogenetic diversity play key roles in modulating communities of hosts and parasitoids and their interactions with each other. These interactions are further structured by the phylogenetic associations between hosts and parasitoids. Moreover, canopy cover partly determined host-parasitoid networks, supporting a recent finding that the structure of host-parasitoid networks is also mediated by changes in microclimate to some extent (Fornoff et al. 2021). These patterns were highly associated with multiple tree diversity metrics (tree taxonomic, phylogenetic and functional diversity), and compositional changes which are key to understanding how host-parasitoid interactions may be impacted by biodiversity loss through trait- and phylogeny-based processes.

Community composition associations

Host species composition was influenced by several factors, including the taxonomic and phylogenetic composition of trees and parasitoids, tree diversity (species richness and MPD), and other biotic factors. The effects of tree diversity and composition on host species composition agree with previous studies where solitary bee and wasp species composition were related to plant community structure (e.g. Loyola and Martins 2008). It seems likely that these results are based on bee linkages to pollen resources and predatory wasp linkages to the diverse of food sources, which may themselves be closely linked to tree species richness (Reitalu et al. 2019, Staab and Schuldt 2020).

We also found that tree MPD, FD, and species composition affect parasitoid species composition, as many studies have found significant relationships between plant and parasitoid diversity in forests, including tree phylogenetic diversity (Staab et al. 2016), functional diversity (Rodriguez et al. 2019), and structural diversity (Schuldt et al. 2019). Similar to predators, parasitoids might also be more active or efficient with increasing tree community dissimilarity due to higher prey resources or lower intraguild parasitism caused by more diverse habitats (Finke and Denno 2002). On the other hand, our results also show that host species composition and parasitoid species composition relate to each other and their phylogenetic compositions, which are structured by tree communities to some extent. This pattern could propagate to the adjacent two higher trophic interactions through both top-down and bottom-up control.

Both host and parasitoid phylogenetic composition were related to tree species composition. This pattern has important implications for cascading effects among trophic levels, in that producer communities could structure a higher trophic level community via an intermediate trophic level. As previous studies usually found weak effects of plants on higher trophic levels (e.g. Cappelli et al. 2022). However, while both host and parasitoid phylogenetic composition was related to tree MPD, only parasitoids responded to tree phylogenetic composition. This may be because there are many caterpillar-hunting wasps in our host communities, and the community composition of caterpillars were usually correlated with tree phylogenetic communities(Wang et al. 2019). Therefore, the prey highly associated with tree phylogenetic composition (e.g. caterpillars) might indirectly determine predatory wasp (host) phylogenetic composition, as recently found for plants-caterpillars-spiders (Chen et al. 2023). This could be further tested by collecting the food directly used by the wasps (caterpillars). For parasitoids, tree phylogenetic composition might drive the process of community assembly through trophic cascades (e.g. from plants to parasitoids via herbivores and host wasps) (Webb et al. 2002, Cavender-Bares et al. 2009). Additionally, parasitoid phylogenetic composition can be influenced by tree structural diversity (e.g. host availability in plots with higher heterogeneity; Schuldt et al. 2019), which can be determined by conserved traits across phylogenies (Webb et al. 2002). The phylogenetic associations between hosts and parasitoids exhibited a nonrandom structure (significant parafit correlation) between the phylogenetic trees of the host and their parasitoids (see also Peralta et al. 2015).

Interestingly, we found that only host phylogenetic composition was affected by parasitoid phylogenetic composition (and not vice-versa). This asymmetry suggests that top-down control (parasitoids to hosts) was stronger than bottom-up control (hosts to parasitoids), supporting prior hypotheses, demonstrating the strong control of parasitoids (Vidal and Murphy 2018). This pattern might reflect disproportionate interactions between hosts and parasitoids, with one trophic level more strongly influencing the other in some instances, but further study is needed.

Moreover, the species and phylogenetic composition of hosts and parasitoids was also related to abiotic factors, especially to canopy cover, which has been considered especially important (Sobek et al. 2009, Fornoff et al. 2021). In future studies, it will be useful to incorporate other, more direct metrics of microclimate, such as local temperature and humidity, to determine the proximal drivers of these microclimatic effects (Ma et al. 2010, Fornoff et al. 2021).

Community-level host-parasitoid networks

Tree community species richness did not significantly influence the diversity of hosts targeted by parasitoids (generality), but caused a significant increase in the diversity of parasitoids per host species (vulnerability). This is likely because niche differentiation often influences network specialization via potential higher resource diversity in plots with higher tree diversity (Lopez-Carretero et al. 2014). For the significant relationship between vulnerability and tree species richness, a potential explanation is that host-parasitoid interactions could be driven through bottom-up effects. Moreover, higher trophic levels will particularly benefit from tree diversity, inducing parasitoid species increases more than host diversity with increasing tree species richness (Guo et al. 2021). These will further increase vulnerability at community level. According to the enemies hypothesis (Root 1973), which posits a positive effects of plant richness on natural enemies, the higher trophic levels in our study (e.g. predators and parasitoids) would benefit from tree diversity and regulate herbivores thereby (Staab and Schuldt 2020). Indeed, previous studies at the same site found that bee parasitoid richness and abundance were positively related to tree species richness, but not their bee hosts (Fornoff et al. 2021, Guo et al. 2021). Because our dataset considered all hosts and reflects an overall pattern of host-parasitoid interactions, the effects of tree species richness on generality might be more complex and difficult to predict, as we found that neither tree species richness nor tree MPD were related to generality. Thus, our results again indicate that top-down control imposed by higher trophic levels is more significant than bottom-up control with an increasing number of plant species.

Linkage density was positively related to tree species richness, supporting the food web theory, which predicts in our case that network complexity (linkage density) depends on the number of plants (Melián and Bascompte 2002). Although trees were not directly included as a trophic level in our networks, potential network complexity increased with tree species richness, likely enabling higher network stability/resistance (Ebeling et al. 2011). For example, a network might be more sensitive to extinctions because of key species loss due to lower linkage density and lower redundancy (e.g. Naeem and Li 1997). However, parasitoid robustness was unexpectedly negatively related to tree species richness. It was expected that higher trophic levels would be more robust, less influenced by perturbations from lower trophic levels, when plant diversity is higher, as more potential interactions at lower trophic levels should theoretically increase redundancy and resilience of connected higher levels (Blüthgen and Klein 2011, Fornoff et al. 2019). Dilution effects may explain this, as plots with higher richness held fewer individuals of a given tree species. If there are strong prey item (caterpillars, grasshoppers, etc.) preferences for one species, there may be fewer or they may be more densely aggregated and less likely to be encountered by parasitoids. This increased stochasticity in parasitoid wasps could benefit hosts by reducing parasitism pressure overall, weakening top-down controls.

Similar to tree species richness, tree MPD was also positively correlated with vulnerability and linkage density, meaning that the mean number of parasitoids per host species and number of links within the host-parasitoid system can also be promoted by tree MPD, in agreement with several recent studies (Pellissier et al. 2013, Staab et al. 2020, Wang et al. 2020). Our results suggest that the specialization and complexity of higher trophic levels can also be affected by plant phylogenetic diversity. This pattern can be traced to the effects of habitat heterogeneity caused by tree species richness and MPD on higher trophic levels via bottom-up control. The effects of tree MPD were consistent with effects of tree species richness on robustness of parasitoids to host loss. This result suggests that higher trophic levels are sensitive to changes in both plant phylogenetic relatedness and general species dissimilarity via trophic interactions, even the hosts are not all directly interacting with plants, bees excluded. Therefore, it may be that stronger linkages would be found when exclusively exploring such plant-herbivore-parasitoid systems.

Interaction evenness was significantly negatively related to canopy cover, further reinforcing an important role of microclimate (likely temperature and humidity; (Sobek et al. 2009, Fornoff et al. 2021). Our results agree with a previous study on ants, where plant-insect interactions were more even with more open canopies (Dáttilo and Dyer 2014). In our case, canopy cover might change hymenopteran species evenness and then further influence interaction evenness. certain host species tended to nest in plots with higher canopy cover, which might decrease the interaction evenness by favoring parasitoids of fewer, more dominant hosts. This pattern would become more significant when more host and parasitoid species are in a plot, given the positive relationship between higher trophic level diversity and canopy cover.

Future prospects

Overall, our study enables new insights into the dynamics of host-parasitoid interactions under varying environmental conditions, an important step toward building a synthetic model for such biodiversity. A key finding was that although parasitoids and hosts respond to tree species richness, top-down control seems predominant in parasite-host interaction, though whether this holds for others antagonistic interactions requires further investigation.

Different trophic levels and functional groups of species responded differently to lower level changes (Fornoff et al. 2021, Guo et al. 2021). This highlights the complexities of building such models and calls for more studies across habitat types and taxa, to test the generality of our findings. Future studies should also consider the role of host/parasitoid functional traits, because they might play a critical role in modifying network structures and ecosystem functioning.

Materials and methods

Study sites design

This study was conducted in the BEF-China biodiversity experiment, which is the largest tree diversity experiment worldwide. The experiment is located in a subtropical forest near Xingangshan, Jiangxi province, south-east China (29°08′–29°11′N, 117°90′–117°93′E). The mean annual temperature is 16.7°C and mean annual precipitation 1821 mm (Yang et al. 2013). The experiment includes two study sites (Site A and Site B), 4 km apart from each other, that were established in 2009 (Site A) and 2010 (Site B) respectively. A total of 566 plots (25.8×25.8 m) were designed on the two sites, and per plot 400 trees were initially planted in 20 rows and 20 columns with a planting distance of 1.29 m. A tree species richness gradient (1, 2, 4, 8, 16 and 24 species) was established at each site, based on a species pool of 40 local, broadleaved tree species (Bruelheide et al. 2014).

For our study, at both sites (site A and site B) eight plots of each tree species richness level (1, 2, 4, 8) were randomly selected, as well as six and two plots of 16 and 24 mixtures. In addition, at site B eight additional monocultures were sampled (Fornoff et al. 2021), resulting 48 plots in Site B (including 16 monocultures, eight plots for each 2, 4, 8 mixtures and six and two plots of 16 and 24 mixtures. In total, 88 study plots were used (40 plots on Site A and 48 plots on Site B, see Fig. S1).

Sampling

We collected trap nests monthly to sample solitary bees and wasps (Staab et al. 2018) in the 88 plots from September to November in 2015 and April to November in 2016, 2018, 2019 and 2020. For each plot, we installed two poles with trap nests (11 m apart from each other and 9 m away from the nearest adjacent plots) along a SW–NE diagonal centrally per plot (following the design of Ebeling et al. 2012). Each pole stood 1.5 m above ground, and each trap nest consisted of two PVC tubes (length: 22 cm × diameter: 12.5 cm) filled with 75 ± 9 (SD) reed internodes of 20 cm length and diameters varying between 0.1 and 2.0 cm (Staab et al. 2014, Fornoff et al. 2021).

Every month, we sampled the reeds with nesting hymenopterans and replaced them with internodes of the same diameter. All the samples were reared in glass test tubes under ambient conditions until specimens hatched. We identified hatched hosts and parasitoids to species or morphospecies (Supplementary Table S1) based on reference specimens (vouchered at the Institute of Zoology, CAS, Beijing). We were interested in the general patterns of host-parasitoid interactions at the community level, so for the analysis we did not distinguish between the two life-history strategies of parasitoids (true parasitoids and kleptoparasitoids, including hymenopteran and dipteran parasitoids) because they both have the same ecological result, death of host brood cells.

DNA extraction and amplification

All specimens were sequenced for a region of the mitochondrial cytochrome c oxidase subunit I (COI) gene (Hebert et al. 2003). We extracted whole-genomic DNA of hosts and parasitoids using DNeasy Blood & Tissue Kits (QIAGEN GmbH, Hilden, Germany), following the manufacturer’s protocols. COI sequences of samples were amplified using universal primer pairs, LCO1490 (GGTCAACAAATCATAAAGATATTGG) as the forward primer and HCO2198 (TAAACTTCAGGGTGACCAAAAAATCA) or HCOout (CCAGGTAAAATTAAAATATAAACTTC) as the reverse primer. We carried out polymerase chain reactions (PCR) in 96-well plates with 30 μl reactions containing 10 μl ddH2O, 15 μl Premix PrimeSTAR HS (TaKaRa), 1 ul of each primer at 10 μM, and 3 μl template genomic DNA using a thermo cycling profile. The PCR procedure as follows: 94℃ for 1 min; 94℃ for 1 min, 45℃ for 1.5 min and 72℃ for 1.5 min, cycle for 5 times; 94℃ for 2 min, 58℃ for 1.5 min and 72℃ for 1 min, cycle for 36 times; 72℃ for 5 min. We performed all PCRs on an Eppendorf Mastercycler gradient, which were then visualized on a 1% agarose gel. Samples with clean single bands were sequenced after PCR purification using BigDye v3.1 on an ABI 3730xl DNA Analyser (Applied Biosystems).

Sequence alignment and phylogenetic analysis

We applied MAFFT (Misawa, Katoh, Kuma, & Miyata, 2002) to align all sequences, then translated the nucleotides into amino acids via MEGA v7.0 (Kumar, Stecher, & Tamura, 2016) to check for the presence of stop codons with manual adjustments. Host and parasitoid sequences were then aligned against the references using a Perl-based DNA barcode aligner (Chesters, 2019).

We employed two strategies to improve the phylogenetic structure of a DNA barcode phylogeny, which demonstrably improve resulting phylogeny-based diversity indices (Macías-Hernández et al. 2020). These include the integration of 1) molecular sequences of the plot data and 2) phylogenetic relationships from other molecular datasets. Integration was achieved following Wang et al. (2020) and Chesters (2020): reference DNA barcodes of Hymenoptera and Diptera were downloaded from the BOLD API (www.boldsystems.org/index.php/API_Public), which were variously processed (e.g. to retain only fully taxonomically labelled barcodes, to remove low quality or mislabeled entries, and to dereplicate to a single exemplar per species), and then aligned (Chesters 2019). A single outgroup was included for which we selected the most appropriate insect order sister to Diptera and Hymenoptera (Misof et al. 2014), a representative of the order Psocoptera (Psocidae, Psocus leidyi). We then constructed a phylogeny of the references and subjects, with references constrained according to the method described earlier (Chesters 2020). A number of backbone topologies were integrated for setting hard and soft constraints, including a transcriptomics-derived topology (Chesters 2020), a mitogenome tree of insects (Chesters 2017), Diptera-specific trees (Wiegmann et al. 2011, Cranston et al. 2012, Ament 2017) and Hymenoptera-specific trees (Peters et al. 2011, Branstetter et al. 2017, Cardinal 2018). The constrained inference was conducted with RaxML version 8 (Stamatakis 2014) under the standard GTRGAMMA DNA model with 24 rate categories. According to the backbone trees used, most taxa present were monophyletic with a notable exception of Crabronidae, for which there is emerging phylogenomic evidence of its polyphyly (Sann et al. 2018).

Tree phylogenetic diversity, functional diversity, and environmental covariates

The phylogenetic diversity of the tree communities was quantified by calculating wood volume-weighted phylogenetic Mean Pairwise Distance (MPD) (Tucker et al. 2017). Tree wood volume was estimated from data on basal area and tree height (Bongers et al. 2021) measured in the center of each plot. Moreover, to represent variations towards the tips of the phylogeny beyond MPD, we additionally calculated Mean Nearest Taxon Distance (MNTD), which is a measure that quantifies the distance between each species and its nearest neighbor on the phylogenetic tree (Webb 2000). Phylogenetic metrics of trees were calculated based on a maximum likelihood phylogenetic tree available for the tree species in our study area (Purschke et al. 2017). Considering that predatory wasps mainly feed on herbivorous caterpillars, we calculated tree functional diversity to test the indirect effects on hymenopteran communities and relevant network indices. Specifically, seven leaf traits were expressed as Rao’s Q (Ricotta and Moretti 2011), including specific leaf area, leaf toughness, leaf dry matter content, leaf carbon content, ratio of leaf carbon to nitrogen, leaf magnesium content, and leaf calcium content. These functional traits were commonly related to higher trophic levels in our study area (Wang et al. 2020, Chen et al. 2023), which are the main food resources of our predatory wasps. All of the traits were measured on pooled samples of sun-exposed leaves of a minimum of five tree individuals per species following standard protocols (Pérez-Harguindeguy et al. 2003).

As our analyses mainly compare community patterns among study plots, we additionally considered potential effects of environmental variation by using plot means of slope, elevation, “eastness” (sine-transformed radian values of aspect), and “northness” (cosine-transformed radian values of aspect) as environmental covariates that characterize the heterogeneity of the study plots. We also accounted for the potential effects of canopy cover at plot level for host-parasitoid interactions, as it can structure hymenopteran communities (Perlík et al. 2023). Canopy cover was calculated as in Fornoff et al. (2021) based on above-nest images.

Statistical analysis

All analyses were conducted in R 4.1.2 with the packages ape, vegan, picante, bipartite, and caper (http://www.R-project.org). Prior to analysis, samples from the five years (2015, 2016, 2018, 2019, and 2020) were pooled at the plot level to discern overall and generalizable effects permeating this system. We excluded three plots with no living trees because of high mortality, resulting in 85 plots in the final analysis.

Composition of trees, hosts and parasitoids

The species and phylogenetic composition of trees, hosts, and parasitoids were quantified at each plot with nonmetric multidimensional scaling (NMDS) analysis based on Morisita-Horn distances. The phylogenetic composition was calculated by mean pairwise distance among the host or parasitoid communities per plot with the R package “picante.” To test the influence of study plot heterogeneity on these relationships, we fitted their standardized values to the ordination on the basis of a regression with the NMDS axis scores (Quinn and Keough 2002). For the analysis, we considered tree species richness, tree functional and phylogenetic diversity, canopy cover, and environmental covariates (elevation, eastness, northness, and slope) as plot characteristics. We assessed the significance of correlations with permutation tests (permutation: n= 999).

Phylogenetic match of hosts and parasitoids

In addition, we used a parafit test (9,999 permutations) with the R package “ape” to test whether the associations were non-random between hosts and parasitoids. The species that were not attacked by parasitoids or failed to generate sequences were excluded from the analyses. For species abundance and composition, see Table S1.

Host-parasitoid interactions

We constructed quantitative host-parasitoid networks with the R package “bipartite” for the level of host and parasitoid communities across for each plot of the two sites. All hosts were grouped because there were too few abundant bee species to separate them out from other hosts. We calculated six indices to quantitatively characterize the structure of the interaction networks: weighted generality (effective number of host species per parasitoid species), weighted vulnerability (Effective number of parasitoid species attacking a host species), robustness (degree of network stability), linkage density (degree of network specialization), and interaction evenness (degree of network evenness). Generality was defined as the weighted mean number of host species per parasitoid species, , with Aj being the number of interactions of parasitoid species j, m the total number of interactions of all species, and Hj the Shannon diversity of interactions of species j. Vulnerability was the weighted mean number of parasitoid species per host species, (Bersier et al. 2002). Robustness was defined as the area under the extinction curve, reflecting the degree of decreases of one trophic level with the elimination of the other trophic levels, here using the robustness index for higher trophic levels (i.e. parasitoids). For linkage density, , we used the realized proportion of possible links between the two trophic levels as the mean number of interactions per species across the entire network (Tylianakis et al. 2007). Interaction evenness was defined as ES = − ∑i ∑j pijlnpij /lnIJ, which is used to describe Shannon’s evenness of network interactions (Dormann et al. 2009). To check whether all network indices significantly differ from chance across all study plots, we used Patefield null models (Dormann et al. 2009) to compare observed indices with simulated values (10,000 times).

Linear mix-effects models

To test the effects of tree species richness, tree phylogenetic, and functional diversity, as well as canopy cover and the other environmental covariates (including slope, elevation, eastness, and northness) on the six network indices, we used linear mixed-effect models. For our analyses, the study sites were considered as random effects, and the others were treated as fixed effects (see above). Given the strong correlation between tree species richness and tree MPD (Pearson’s r = 0.74, p<0.01), we excluded tree MPD in the models where tree species richness was a predictor. To evaluate the potential effects caused by tree MPD, we also ran alternative models where tree species richness was replaced with tree MPD. We simplified all models by gradually removing non-significant factors to obtain the most parsimonious model with the lowest AICc. To ensure that the analyses were not strongly affected by multicollinearity, the correlations among all predictors were tested (Fig. S2), and variance inflation factors (VIF) of our statistical models were checked.

eLife Assessment

The valuable work by authors improves our understanding on the effects of tree diversity on host-parasitoid communities of insects in forests in an experimental setting. Most of the analyses used are solid, but some of the conclusions seem a bit too strong and stretched.

Reviewer #1 (Public Review):

Summary:

The authors analyzed how biotic and abiotic factors impact antagonistic host-parasitoid interaction systems in a large BEF experiment. They found the linkage between the tree community and host-parasitoid community from the perspective of the multi-dimensionality of biodiversity. Their results revealed that the structure of the tree community (habitat) and canopy cover influence host-parasitoid compositions and their interaction pattern. This interaction pattern is also determined by phylogenetic associations among species. This paper provides a nice framework for detecting the determinants of network topological structures.

Strengths:

This study was conducted using a five-year sampling in a well-designed BEF experiment. The effects of the multi-dimensional diversity of tree communities have been well explained in a forest ecosystem with an antagonistic host-parasitoid interaction.

The network analysis has been well conducted. The combination of phylogenetic analysis and network analysis is uncommon among similar studies, especially for studies of trophic cascades. Still, this study has discussed the effect of phylogenetic features on interacting networks in depth.

Weaknesses:

(1) The authors should examine species and interaction completeness in this study to confirm that their sampling efforts are sufficient.

(2) The authors only used Rao's Q to assess the functional diversity of tree communities. However, multiple metrics of functional diversity exist (e.g., functional evenness, functional dispersion, and functional divergence). It is better to check the results from other metrics and confirm whether these results further support the authors' results.

(3) The authors did not elaborate on which extinction sequence was used in robustness analysis. The authors should consider interaction abundance in calculating robustness. In this case, the author may use another null model for binary networks to get random distributions.

(4) The causal relationship between host and parasitoid communities is unclear. Normally, it is easy to understand that host community composition (low trophic level) could influence parasitoid community composition (high trophic level). I suggest using the 'correlation' between host and parasitoid communities unless there is strong evidence of causation.

Reviewer #2 (Public Review):

Summary:

In their manuscript, Multi-dimensionality of tree communities structure host-parasitoid networks and their phylogenetic composition, Wang et al. examine the effects of tree diversity and environmental variables on communities of reed-nesting insects and their parasitoids. Additionally, they look for the correlations in community composition and network properties of the two interacting insect guilds. They use a data set collected in a subtropical tree biodiversity experiment over five years of sampling. The authors find that the tree species, functional, and phylogenetic diversity as well as some of the environmental factors have varying impacts on both host and parasitoid communities. Additionally, the communities of the host and parasitoid showed correlations in their structures. Also, the network metrices of the host-parasitoid network showed patterns against environmental variables.

Strengths:

The main strength of the manuscript lies in the massive long-term data set collected on host-parasitoid interactions. The data provides interesting opportunities to advance our knowledge on the effects of environmental diversity (tree diversity) on the network and community structure of insect hosts and their parasitoids in a relatively poorly known system.

Weaknesses:

To me, there are no major issues regarding the manuscript, though sometimes I disagree with the interpretation of the results and some of the conclusions might be too far-fetched given the analyses and the results (namely the top-down control in the system). Additionally, the methods section (especially statistics) was lacking some details, but I would not consider it too concerning. Sometimes, the logic of the text could be improved to better support the studied hypotheses throughout the text. Also, the results section cannot be understood as a stand-alone without reading the methods first. The study design and the rationale of the analyses should be described somewhere in the intro or presented with the results.

Author response:

The following is the authors’ response to the original reviews.

It would be great if the authors could add clarification about the NMDS analyses and the associated results (Fig. 1, Table 1 and Tables S2-4). The overall aim of these analyses was to see how plot characteristics (e.g. canopy cover) and composition of one taxonomic group were related to the composition of another taxonomic group. The authors quantified species composition by two axes from NMDS. (1) This analysis may yield an interpretation problem: if we only find one of the axes, but not the other, was significantly related to one variable, it would be difficult to determine whether that specific variable is important to the species composition because the composition is co-determined by two axes. (2) It is unclear how the authors did the correlation analyses for Tables S2-4. If correlation coefficients were presented in these tables, then these coefficients should be the same or very similar if we switch the positions of y vs. x. That is, the correlation between host vs. parasite phylogenetic composition would be very close to the correlation between parasite vs. phylogenetic composition, but not as the author found that these two relationships were quite different, leading to the interpretation of bottom-up or top-down processes. It is also unclear which correlation coefficient was significant or not because only one P value was provided per row in these tables. (3) In addition to the issues of multiple axes (point 1), NMDS axes simply define the relative positions of the objects in multi-dimensional space, but not the actual dissimilarities. Other methods, such as generalized dissimilarity modeling, redundancy analysis and MANOVA, can be better alternatives.

Thank you for the thorough and constructive review. We have taken the concerns and questions raised by the editors and reviewers into account and provided clarification about the NMDS analyses as well as additional analyses to confirm our results. First, we have now added a brief explanation in the manuscript regarding the interpretation of the two NMDS axes and how they relate to species composition. Specifically, we clarified that while NMDS defines the relative positions of objects in multi-dimensional space, the two axes together provide a more comprehensive representation of the community composition, which is not solely determined by either axis independently. Second, we acknowledge that alternative approaches could help further strengthen our conclusions. To address this, we incorporated Mantel tests and PERMANOVA (with ‘adonis2’) as additional validation methods. These analyses allowed us to summarize compositional patterns while testing our hypotheses within the framework of the plot characteristics and taxonomic relationships. We have added these analyses and their results in the manuscript to reinforce our findings.

In methods: L478-481 “To strengthen the robustness of our findings based on NMDS, we further validated the results using Mantel test and PERMANOVA (with ‘adonis2’) for correlation between communities and relationships between communities and environmental variables.”

L469-475 “NMDS was used to summarize the variation in species composition across plots. The two axes extracted from the NMDS represent gradients in community composition, where each axis reflects a subset of the compositional variation. These axes should not be interpreted in isolation, as the overall species composition is co-determined by their combined variation. For clarity, results were interpreted based on the relationships of variables with the compositional gradients captured by both axes together."

In results: L172-177 “The PERMANOVA analysis also highlighted the important role of canopy cover for host and parasitoid community (Table S6-9). The Mantel test revealed a consistent pattern with the NMDS analysis, highlighting a pronounced relationship between the species composition of hosts and parasitoids (Table S10). However, the correlation between the phylogenetic composition of hosts and parasitoids was not significant.”

In discussion: L257-261 “However, this significant pattern was observed only in the NMDS analysis and not in the Mantel test, suggesting that the non-random interactions between hosts and parasitoids could not be simply predicted by their community similarity and associations between the phylogenetic composition of hosts and parasitoids are more complex and require further investigation in the future.”

-- One additional minor point: "site" would be better set as a fixed rather than random term in the linear mixed-effects models, because the site number (2) is too small to make a proper estimate of random component.

Now we treated “site” as a fixed factor in our models, interacting with tree species richness/tree MPD and tree functional diversity to reflect the variation of spatial and tree composition between the two sites. We found the main results did not change, as both sites showed consistent patterns for effects of tree richness/MPD on network metrics, which is more pronounced in one site.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors analyzed how biotic and abiotic factors impact antagonistic host-parasitoid interaction systems in a large BEF experiment. They found the linkage between the tree community and host-parasitoid community from the perspective of the multi-dimensionality of biodiversity. Their results revealed that the structure of the tree community (habitat) and canopy cover influence host-parasitoid compositions and their interaction pattern. This interaction pattern is also determined by phylogenetic associations among species. This paper provides a nice framework for detecting the determinants of network topological structures.

Strengths:

This study was conducted using a five-year sampling in a well-designed BEF experiment. The effects of the multi-dimensional diversity of tree communities have been well explained in a forest ecosystem with an antagonistic host-parasitoid interaction.

The network analysis has been well conducted. The combination of phylogenetic analysis and network analysis is uncommon among similar studies, especially for studies of trophic cascades. Still, this study has discussed the effect of phylogenetic features on interacting networks in depth.

Weaknesses:

(1) The authors should examine species and interaction completeness in this study to confirm that their sampling efforts are sufficient.

(2) The authors only used Rao's Q to assess the functional diversity of tree communities. However, multiple metrics of functional diversity exist (e.g., functional evenness, functional dispersion, and functional divergence). It is better to check the results from other metrics and confirm whether these results further support the authors' results.

(3) The authors did not elaborate on which extinction sequence was used in robustness analysis. The authors should consider interaction abundance in calculating robustness. In this case, the author may use another null model for binary networks to get random distributions.

(4) The causal relationship between host and parasitoid communities is unclear. Normally, it is easy to understand that host community composition (low trophic level) could influence parasitoid community composition (high trophic level). I suggest using the 'correlation' between host and parasitoid communities unless there is strong evidence of causation.

Thank you very much for your thoughtful and constructive review of our manuscript. We have carefully addressed your comments and made several revisions to improve the clarity and robustness of our work.1) We appreciate your suggestion regarding species and interaction completeness. To confirm that our sampling efforts were sufficient, we have now included a figure (Fig. S1) showing the species accumulation curve and the coverage of interactions in our study. This ensures that the data collected provide a comprehensive representation of the system. 2) Regarding the use of only Rao’s Q to assess functional diversity, we acknowledge that multiple metrics of functional diversity exist. However, due to the large number of predictors in our analysis, we decided to streamline our approach and focus on Rao’s Q as it provides a robust measure for our research objectives. We have discussed this decision in the revised manuscript and clarified that, while additional metrics could be informative, we believe Rao’s Q sufficiently captures the key aspects of functional diversity in our study. 3) We have elaborated on the robustness analysis and the null model used in our study. Specifically, we now clarified which extinction sequence (random extinction) was used in our manuscript, and explained interaction abundance was incorporated into the robustness calculations (networklevel function, weighted=TURE; see L506). 4) We have revised the text to clarify the relationship between host and parasitoid communities. As you correctly pointed out, while it is intuitive that host community composition influences parasitoid community composition, we have reframed our analysis to emphasize the correlation between the two communities rather than implying causation without strong evidence. We have revised the manuscript to reflect this distinction.

Reviewer #2 (Public Review):

Summary:

In their manuscript, Multi-dimensionality of tree communities structure host-parasitoid networks and their phylogenetic composition, Wang et al. examine the effects of tree diversity and environmental variables on communities of reed-nesting insects and their parasitoids. Additionally, they look for the correlations in community composition and network properties of the two interacting insect guilds. They use a data set collected in a subtropical tree biodiversity experiment over five years of sampling. The authors find that the tree species, functional, and phylogenetic diversity as well as some of the environmental factors have varying impacts on both host and parasitoid communities. Additionally, the communities of the host and parasitoid showed correlations in their structures. Also, the network metrices of the host-parasitoid network showed patterns against environmental variables.

Strengths:

The main strength of the manuscript lies in the massive long-term data set collected on host-parasitoid interactions. The data provides interesting opportunities to advance our knowledge on the effects of environmental diversity (tree diversity) on the network and community structure of insect hosts and their parasitoids in a relatively poorly known system.

Weaknesses:

To me, there are no major issues regarding the manuscript, though sometimes I disagree with the interpretation of the results and some of the conclusions might be too far-fetched given the analyses and the results (namely the top-down control in the system). Additionally, the methods section (especially statistics) was lacking some details, but I would not consider it too concerning. Sometimes, the logic of the text could be improved to better support the studied hypotheses throughout the text. Also, the results section cannot be understood as a stand-alone without reading the methods first. The study design and the rationale of the analyses should be described somewhere in the intro or presented with the results.

Thank you very much for your valuable comments and suggestions on our manuscript! We appreciate your feedback and have made revisions accordingly. Specifically, we have rephrased the interpretation of the results and conclusions to better align with the analyses and avoid overstatements, particularly concerning the top-down control in the system. In addition, we have expanded the methods section by providing more details, especially regarding the statistical approaches, to address the points you raised. To enhance the clarity of the manuscript, we have also ensured that the logic of the text better supports the hypotheses throughout. Please see our point-by-point responses below for additional clarifications.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Line 120: "... and large ecosystems susceptible to global change (add citation here)": Citation(s)?

Now we provided the missed citations.

Line 141: Add sampling completeness information.

Now we provide a new figure about sampling completeness (Fig. S1) in the supplementary materials, showing the adequate sampling effort for our study.

Line 151: use more metrics in the evaluation of functional diversity

We used tree functional diversity Rao’s Q, which is an integrated and wildly used metric to represent functional dissimilarity of trees. As our study focus on multiple diversity indices of trees, it would be better to do not pay more attention to one type of diversity. Thank you for your suggestion!

Line 164: host vulnerability. Although generality and vulnerability are commonly used in network analysis, it is better to link these metrics with the trophic level, like the 'host vulnerability' you used. Thus, you can use 'parasitoid generality' instead of 'generality'.

Thanks for your suggestion. Now the metrics were labeled with the trophic levels in the full text.

Line 169: two'.'

Corrected.

Line 173: 'parasitoid robustness' Or 'robustness of parasitoids'?

Now changed it to ‘robustness of parasitoid’.

Lines 173, 468: For the robustness estimations, maybe use null model for binary networks to get random distributions?

Thanks for the suggestion. Actually, we have used Patefield null models to compare the randomized robustness and observed, helping to assess whether the robustness of the observed network is significantly different compared to expected by chance. All robustness indices across plots were significantly different from a random distribution, See results section L197-201.

Line 184: modulating interacting communities of hosts and parasitoids.

Changed accordingly.

Line 186: determined host-parasitoid interaction patterns

Changed accordingly.

Line 191: Biodiversity loss in this study refers to low trophic levels.

Now we clarified this point.

Line 190: understand

Changed accordingly.

Lines 215-216: Reorganize these sentences

Line 227: indirectly influenced by...

Changed accordingly.

Line 238: Be more specific. Which type of further study?

Rephased it more specific.

Lines 297-299: rewrite this sentence to make it more transparent.

Now we rewrote the sentence accordingly.

Line 302: Certain

Changed accordingly.

Line 453: effective

Changed accordingly.

Finally, the authors should check the text carefully to avoid grammatical errors.

Thanks, now we have checked the full text to avoid grammatical errors.

Reviewer #2 (Recommendations For The Authors):

I feel that the authors have very interesting data and have a solid set of analyses. I do not have major issues regarding the manuscript, though sometimes I disagree with the interpretation of the results and some of the conclusions might be too far-fetched given the analyses and the results. Additionally, the methods section (especially statistics) was lacking some details, but I would not consider it too concerning at this point.

I feel that the largest caveat of the manuscript remains in the representation of the rationale of the study. I felt the introduction could be more concise and be better focused to back up the study questions and hypotheses. Many times, the sentences were too vague and unspecific, and thus, it was difficult to understand what was meant to be said. The authors could mention something more about how community composition of hosts and parasitoids are expected to change with the studied experimental design regarding the metrices you mention in the introduction (stronger hypotheses). The results section cannot be understood as a stand-alone without reading the methods first. The study design and the rationale of the analyses must be described somewhere in the intro or results, if the journal/authors want to keep the methods last structure. Also, the results and discussion could be more focused around the hypotheses. Naturally, these things can be easily fixed.

I also disagree with the interpretation of results finding top-down control in the system (it might well be there, but I do not think that the current methods and tests are suitable in finding it). First, the used methodology cannot distinguish parasitoids if the hosts are not there and the probability to detect parasitoid likely depends on the abundance of the host. Thus, the top-down regulation is difficult to prove (is it the parasitoids that have driven the host population down). Secondly, I would be hesitant to say anything about the top-down and bottom-up control in the systems as the data in the manuscript is pooled across five years while the top-down/bottom-up regulation in insect systems usually spans only one season/generation in time (much shorter than five years). Consequently, the analyses are comparing the communities of species that some of most likely do not co-exist (they were found in the same space but not during the same time). Luckily, the top-down/bottom-up effects could potentially be explored by using separately the time steps of the now pooled community data: e.g., does the population of the host decrease in t if the parasitoids are abundant in t-1? There are also other statistical tests to explore these patterns.

In the manuscript "Phylogenetic composition" refers to Mean Pairwise Distance. I would use "phylogenetic diversity" instead throughout the text. Also, to my understanding, in trees both "phylogenetic composition" and "phylogenetic diversity" are used even though based on their descriptions, they are the same.

Detailed comments:

Punctuation needs to be checked and edited at some point (I think copy-pasting had left things in the wrong places). Please check that "-" instead of "-" is used in host-parasitoid.

1-2 The title is not very matching with the content. "Multi-dimensionality" is not mentioned in the text. "phylogenetic composition" -> "phylogenetic diversity"

We didn’t find the role of functional diversity of trees in host-parasitoid interactions, but we still have tree richness and phylogenetic diversity. I also disagree with that using phylogenetic diversity to replace phylogenetic composition, because diversity highlights higher or lower phylogenetic distance among communities, while the later highlights the phylogenetic dissimilarity across communities.

53-57 This sentence is quite vague and because of it, difficult to follow. Consider rephrasing and avoiding unspecified terms such as "tree identity", "genetic diversity", and "overall community composition of higher trophic levels" (at least, I was not sure what taxa/level you meant with them).

Rephased.

L58-61 “Especially, we lack a comprehensive understanding of the ways that biotic factors, including plant richness, overall community phylogenetic and functional composition of consumers, and abiotic factors such as microclimate, determining host–parasitoid network structure and host–parasitoid community dynamics.”

56 I would remove "interact" as no interactions were tested.

Removed accordingly.

59-60 This needs rephrasing. I feel "taxonomic and phylogenetic composition should be just "species composition". To better match, what was done: "taxonomic, phylogenetic, and network composition of both host and parasitoid communities" -> "species and phylogenetic diversity of both host and parasitoid communities and the composition their interaction networks"

Changed accordingly.

62 Remove "tree composition".

Done.

62 Replace "taxonomic" with "species". Throughout the text.

Done.

63-64 "Generally, top-down control was stronger than bottom-up control via phylogenetic association between hosts and parasitoids" I disagree, see my comments elsewhere.

Now we rephased the sentence.

L68-70 “Generally, phylogenetic associations between hosts and parasitoids reflect non-randomly structured interactions between phylogenetic trees of hosts and parasitoids.”

68 "habitat structure and heterogeneity" This is too strong and general of a statement based on the results. You did not really measure habitat structure or heterogeneity.

Now we rephased the statement to avoid strong and general description.

L71-73 “Our study indicates that the composition of higher trophic levels and corresponding interaction networks are determined by plant diversity and canopy cover especially via trophic phylogenetic links in species-rich ecosystems.”

69 Specify "phylogenetic links". Trophic links?

Specified to “trophic phylogenetic links”.

75-77 The sentence is a bit difficult to follow. Consider rephrasing.

Now we rephased it.

L79-82 “Changes in network structure of higher trophic levels usually coincide with variations in their diversity and community, which could be in turn affected by the changes in producers via trophic cascades”

76 Be more specific about what you mean by "community of trophic levels".

Specified to “community composition”.

79 Remove "basal changes of", it only makes the sentence heavier.

Done.

81 What is "species codependence"?

We sim to describe the species co-occurrence depending on their closely relationships. For clarity, now we changed to “species coexistence”

82 What do you mean by "complex dynamics"?

Rephased to “mechanisms on dynamics of networks”.

83 onward: I would not focus so much on top-down/bottom-up as I feel that your current analyses cannot really say anything too strong about these causalities but are rather correlative.

Thanks, we now removed the relevant contents from the discussion. However, we kept one sentence in the Introduction, because it should be highlighted to make reviewers aware of this (the other text on about this were removed).

89 Remove "environmental".

Done.

90 Specify what you mean by "these forces".

Done.

98-99 I have difficulties following the logic here "potential specialization of their hosts may cascade up to impact the parasitoids' presence or absence". Consider rephrasing.

Now we rephased it.

L101-102 “…and their host fluctuations may cascade up to impact the parasitoids’ presence or absence.”

100 Be more specific with "habitat-level changes".

Specified to “community-level changes”

100 I do not see why host-parasitoid systems would be ideal to study "species interactions". There are much simpler and easier systems available.

Changed to “… one of ideal…”

101-103 "influence of" on what?

Now we rephased the sentence.

L104-105 “Previous studies mainly focused on the influence of abiotic factors on host-parasitoid interactions”

104 Be more specific in "the role of multiple components of plant diversity".

Now we specified "the role of multiple components of plant diversity".

L107-108 “…the role of multiple components of plant diversity (i.e. taxonomic, functional and phylogenetic diversity)…”

106 "diversity associations" of what?

“diversity associations between host and parasitoids”.

108 Specify the "direct and indirect effects".

Now we specified it to “…direct and indirect effects (i.e. one pathway and more pathways via other variables)…”

110-113 A bit heavy sentence to follow. Consider rephrasing.

Now we rephased the sentence to make it more readable.

114 Give an example of "phylogenetic dependences".

Done. Phylogenetic dependences (e.g. phylogenetic diversity)

117 Move the "e.g. taxonomic, phylogenetic, functional" within brackets in 117 after "dimensions of biodiversity".

Done.

120 "(add citation here)" Yes please!

Done.

120-121 Specify "such relationships".

Done. Specified to “multiple dimensions of biodiversity”

128-130 This is difficult to follow. Please rephrase.

Now we rephased the sentence.

L135-137 “We aimed to discern the primary components of the diversity and composition of tree communities that affect higher trophic level interactions via quantifying the strength and complexity of associations between hosts and parasitoid.”

131-132 Remove "phylogenetic and". It is redundant to phylogenetic diversity.

Done.

128 Tested robustness does not really capture "stability of associations".

Yes, we agree. Now we rephased the sentence and exclude the “stability” description.

133 Specify "phylogenetic processes".

Now we specified “phylogenetic processes”.

L140-141 “…especially via phylogenetic processes (e.g. lineages of trophic levels diverge and evolve over time)…”

141 I would like to have more details on the data set somewhere in the results. How many individuals and species were found in each plot (on average)? Was there a lot of temporal variation (e.g. between the seasons)? On how many sites were the insect species found?

Thanks for your suggestion. Now we provide more details on the data set in the results (L153-156), including mean values of individuals and species in each plot. However, the temporal variation should be studied for another relative independent topic, as our study focus on the general patter of interactions between hosts and parasitoids. Therefore, we would not put more information on temporal changes to make readers get lost in the text.

153-156 “Among them, we found 56 host species (12 bees and 44 wasps, mean abundance and richness are 400.05 and 45.14, respectively, for each plot) and 50 parasitoid species (38 Hymenoptera and 12 Diptera, mean abundance and richness are 14.07 and 9.05, respectively, for each plot).”

149 tree -> trees

Done.

149 Should there read also some else than "NMDS scores"?

Thanks! Now we provided more details about “NMDS scores”.

L161-162 “(NMDS axis scores; i.e. preserving the rank order of pairwise dissimilarities between samples)”

149 You could mention the amount of variation explained by the first two axes of the NMDSs. Now it is difficult to estimate how much the models actually explain.

Thanks for your comments! However, we could not directly provide the explanatory power of the two axes, because NMDS is based on rank-order distances rather than linear relationships like in PCA. However, the goodness of fit for the NMDS solution is typically evaluated using the stress value. We provide the stress value in the figure caption.

150 "tree MPD" is mentioned for the first time. Spell it out.

Done.

150 Explain "eastness".

Done.

L163-164 “…eastness (sine-transformed radian values of aspect) )”

151 How was "tree functional diversity" quantified?

Please see methods. L437-L438.

160 Specify that you talk about phylogenetic compositions of the host and parasitoid communities here.

We would keep it refined here, keeping consistent with species composition here. Phylogenetic composition just represents the dissimilarities of phylogenetic linages within a community.

161 Describe "parafit" test here when first mentioned.

Done, see methods L485-487.

182 Keep on referring to tables and figures in the discussion! Also, more clearly discuss your hypotheses. There are lots of discussions on top-down/bottom-up control. It could be good to form a hypothesis on them and predict what kind of patterns would suggest either one and what would you expect to find regarding them.

Now we referred figures and tables in the discussion. As the contents on top-down and bottom-up control were not fit very well with our study (as also suggested by reviewers), so we rephased the discussion and also clearly discuss our hypotheses in the discussion. See L218, L226, and L237 etc.

186 "partly determined host-parasitoid networks" Be more specific.

Done.

L206-207 “…partly determined host-parasitoid network indices, including vulnerability, linkage density, and interaction evenness.”

195 Tell what you mean by "other biotic factors".

Specified it: “…other biotic factors such as elevation and slope…”

197-198 "It seems likely that these results are based on bee linkages to pollen resources" I would be hesitant to conclude this as the bees most likely forage way beyond the borders of the 30m by 30m study plots.

Thanks for your concern about this problem. While it is true that bees can forage beyond 30 x 30m, the study focuses on their nesting behavior and activity within this defined area, rather than their entire foraging range. Existing literature shows bees often forage locally when resources are available (e.g. Ebeling et al., 2012 Oecologia; Guo et al., year, Basic and Applied Ecology). Therefore, we are confident that this pattern could be associated with the resources around the trap nests.

223 "This could be further tested by collecting the food directly used by the wasps (caterpillars)" A bit unnecessary addition.

Thanks for your suggestion. Yes, this definitely is a good point, but currently we don’t have enough data of caterpillars, but we will follow this in the future.

232-238 I disagree with the authors on the interpretation of the causality of the results here. I think that the community of parasitoids simply indicates which host species are available, while the host community does not have an as strong effect on parasitoid community as parasitoids do not utilise the whole species pool of the hosts. (Presence of parasitoid tells that the host is around while the presence of the host does not necessarily tell about the presence of the parasitoid.) To me, this would rather indicate a bottom-up than top-down regulation. Similar patterns are also visible in species communities of hosts and parasites.

Thank you for your suggestion. We agree with you that parasitoids are more depended on hosts, as host could not be always attacked by parasitoids. Now we rephased our explanation to follow this argument.

L254-256 “Such pattern could be further confirmed by the significant association between host phylogenetic composition and parasitoid phylogenetic composition (Fig. 1c), which suggested that their interactions are phylogenetically structured to some extent.”

247-266 The logic in this section is difficult to follow. Try rephrasing.

Now we rephased the section for a clearer logic.

L270-287 “Tree community species richness did not significantly influence the diversity of hosts targeted by parasitoids (parasitoid generality), but caused a significant increase in the diversity of parasitoids per host species (host vulnerability) (Fig. 3a; Table 2). This is likely because niche differentiation often influences network specialization via potential higher resource diversity in plots with higher tree diversity (Lopez-Carretero et al. 2014). Such positive relationship between host vulnerability and tree species richness suggested that host-parasitoid interactions could be driven through bottom-up effects via benefit from tree diversity. For example, parasitoid species increases more than host diversity with increasing tree species richness (Guo et al. 2021), resulting increasing of host vulnerability at community level. According to the enemies hypothesis (Root 1973), which posits a positive effects of plant richness on natural enemies, the higher trophic levels in our study (e.g. predators and parasitoids) would benefit from tree diversity and regulate herbivores thereby (Staab and Schuldt 2020). Indeed, previous studies at the same site found that bee parasitoid richness and abundance were positively related to tree species richness, but not their bee hosts (Fornoff et al. 2021, Guo et al. 2021). Because our dataset considered all hosts and reflects an overall pattern of host-parasitoid interactions, the effects of tree species richness on parasitoid generality might be more complex and difficult to predict, as we found that neither tree species richness nor tree MPD were related to parasitoid generality.”

249 "This is likely because niche differentiation often influences network specialization via potential higher resource diversity in plots with higher tree diversity" This is a bit contradicting your vulnerability results as niche differentiation should increase specialization and diversity and specialization should decrease vulnerability (less host per parasitoid).

Thanks! We understand that the concepts of “generality” and “vulnerability” can be a bit confusing. To clarify, “fewer hosts per parasitoid” actually corresponds to lower generality at the community level.

332-337 How did you select the species growing on your plots? Or was only species number considered? What was the pool of tree species growing on the selected plots? Was the selection similar at both sites?

Now we provided more information on the experiment design.

L354-356 “The species pools of the two plots are nonoverlapping (16 species for each site). The composition of tree species within the study plots is based on a “broken-stick” design (see Bruelheide et al. 2014).”

342 Remove "centrally per plot"?

Done.

346-347 Was the selection of different reed diameters similar in all the plots?

Diameters and the relative distribution of diameters was similar in all trap nests.

399 & 432 Are "phylogenetic diversity of the tree communities" and "phylogenetic composition of trees" the same? They are both described as mean pairwise distance.

These two are actually different, as we use this to distinguish the phylogenetic diversity with communities and rank order of dissimilarities between tree communities. Here, the phylogenetic diversity of the tree communities is mean pairwise phylogenetic distance of species for tree communities. Tree phylogenetic composition is the rank order of pairwise dissimilarities between tree communities based on NMDS.

400 Do you think that MPD makes any sense with the monocultures (value is always 0)? Does this have a potential to bias your analyses and result?

We agree your point. However, we do not think that this is a major problem in the analyses. We followed the experimental design and considered low phylogenetic relatedness of tree species in a plot (Likewise in monocultures, the tree species richness is always 1).

402-405 MNTD is not mentioned before or after this. Consider removing this section.

We tested the potential effects of MNTD in our models. Now we mentioned it in our results.

L194-195 “Tree mean nearest taxon distance (MNTD) was unrelated to any network indices.”

405 "Phylogenetic metrics of trees" Which ones?

Both tree MPD and MNTD. Now we have noted it in the manuscript. (L432)

410 Further details on "Rao's Q" and how the functional diversity of the communities was calculated are needed.

Now more details were provided.

L435-438 “Specifically, seven leaf traits were used for calculation of tree functional diversity, which was calculated as the mean pairwise distance in trait values among tree species, weighted by tree wood volume, and expressed as Rao's Q”

413 Specify "higher trophic levels".

Now we specified the trophic levels.

L440-441 “…higher trophic levels in our study area, such as herbivores and predators”

417-424 What about the position of the plots within study sites? Is there potential for edge effects (e.g. bees finding easier the trap nest close to the edge of the experimental forest)? Were there any differences between the two sites? What is the elevation range of the plots?

Thanks for concerning the details of our study. First, all the plots were randomly distributed within the study sites (see Fig. S2). Admittedly, there are several plots are located in the edges of the site. However, we did not consider the potential edge effects in our analysis. Of course, this will be a good point in our future studies. Moreover, the biggest difference between the two is the non-overlapping tree species pool, and the two study sites are apart from 5 km in the same town. Finally, there is not too distinct elevation gradient across the plots (112 m - 260 m).

432-434 "The species and phylogenetic composition of trees, hosts, and parasitoids were quantified at each plot with nonmetric multidimensional scaling (NMDS) analysis based on Morisita-Horn distances" This section needs to be more specific and detailed. Did you do the NMDS separately for each plot as suggested in the text?

We provided more details of the section.

L462-465 “The minimum number of required dimensions in the NMDS based on the reduction in stress value was determined in the analysis (k = 2 in our case). We centred the results to acquire maximum variance on the first dimension, and used the principal components rotation in the analysis.”

435 Specify how picante was used (function and arguments)!

Now we specified the function.

L465-467 “The phylogenetic composition was calculated by mean pairwise distance among the host or parasitoid communities per plot with the R package “picante” with ‘mpd’ function.”

436 "standardized values" Of what? How was the standardisation done?

Now we citied a supplementary table (Table S2) to specify it (see L469). For the standardization, we used ‘scale’ function in R, which standardizes data by centering and scaling data. Specifically, it subtracts the mean and divides by the standard deviation for each variable.

443 Provide more details on parafit.

Actually, we have provided the reason why we use the parafit test and the usage.

L483-486 “We used a parafit test (9,999 permutations) with the R package “ape” to test whether the associations were non-random between hosts and parasitoids. This is widely used to assess host-parasite co-phylogeny by analyzing the congruence between host and parasite phylogenies using a distance-based matrix approach.”

449-451 Rephrase the sentence.

Rephased.

L490-491 “We constructed quantitative host-parasitoid networks at community level with the R package “bipartite” for each plot of the two sites.”

451 "six" Should this be five?

Yes, should be five, thanks.

470-481 What package and function were used for the LMMs?

As we now used linear models, we do no longer use a R package for LMMs.

470 "mix" -> mixed

Changed to linear models.

472 "six" Should this be five?

Again, we changed it to five.

479-481 How did you treat the variables from the two different sites when testing for the correlations to avoid two geographic clusters of data points?

Now we considered the two study sites as fixed factor in our linear models. Moreover, tree-based variables were additionally included as interaction terms with the study sites.

501 "mix" -> mixed

Changed to linear models.

The panel selection for figures 3 and 4 seems random. Justify it!

Thank you. To avoid including too many figures in the main text, which could potentially confuse readers, we have selected the key results that are of primary interest. The remaining figures are provided in the appendix for reference.

533 "Note that axes are on a log scale for tree species richness." Why the log-scale if the analyses were performed with linear fit? Also, the drawn regression lines do not match the model description (non-linear, while a linear model is described in the text). The models should probably be described in more detail.

We used log-transformed to promote the normality of the data. The drawn regression lines are linear lines, which fit our models.

539 "Values were adjusted for covariates of the final regression model." How?

We used residual plot to directly visualizes the relationship between the predictor and the response variable with the fitted regression line, making it easier to assess the model's fit.

Fig. S4 text does not match the figure.

Thanks! We now deleted the unmatched text in the figure.

Version 2

Abstract

Abstract

Environmental factors can influence ecological networks, but these effects are poorly understood in the realm of the phylogeny of host-parasitoid interactions. Especially, we lack a comprehensive understanding of the ways that biotic factors, including plant species richness, overall community phylogenetic and functional composition of consumers, and abiotic factors such as microclimate, determining host–parasitoid network structure and host–parasitoid community dynamics. To address this, we leveraged a five-year dataset of trap-nesting bees and wasps and their parasitoids collected in a highly-controlled, large-scale subtropical tree biodiversity experiment. We tested for effects of tree species richness, tree phylogenetic and functional diversity, and species and phylogenetic composition on species and phylogenetic diversity of both host and parasitoid communities and the composition of their interaction networks. We show that multiple components of tree diversity and canopy cover impacted both, species and phylogenetic composition of hosts and parasitoids. Generally, phylogenetic associations between hosts and parasitoids reflected non-randomly structured interactions between phylogenetic trees of hosts and parasitoids. Further, host-parasitoid network structure was influenced by tree species richness, tree phylogenetic diversity, and canopy cover. Our study indicates that the composition of higher trophic levels and corresponding interaction networks are determined by plant diversity and canopy cover especially via trophic links in species-rich ecosystems.

Introduction

Understanding the ecological consequences of biodiversity loss is an increasingly important task in ecology, given the ongoing biodiversity crisis (Isbell et al. 2022). Representing the interdependencies among organisms, ecological networks reflect whether and how species interact with each other across trophic levels, playing an indispensable role in assessing ecosystem stability and integrity (De Ruiter et al. 1995, Harvey et al. 2017). Changes in network structure of higher trophic levels usually coincide with variations in their diversity and community composition, which could be in turn affected by the changes in producers via trophic cascades (Barnes et al. 2018, Gonzalez et al. 2020). However, we still lack a generalizable framework for how these networks and especially their phylogenetic interdependencies respond to biodiversity loss in ecosystems, such as changes in the tree diversity of forests (Tylianakis et al. 2008, Grossman et al. 2018). To better understand species coexistence and its role for biodiversity conservation, we must further study the mechanisms on dynamics of networks from multiple perspectives (Tittensor et al. 2014, Brondizio et al. 2019). Interactions can be viewed from a top-down or bottom-up perspective. Previous studies have shown there might be asymmetric effects across trophic levels (Vidal and Murphy 2018), with both factors shaping multitrophic communities together(Hunter et al. 1992). Moreover, the diversity and community of higher trophic levels could be also driven by microclimate, which could be determined by plant structuring, such as canopy cover (Fornoff et al. 2021, Perlík et al. 2023). Understanding how canopy cover responds to changing biodiversity is therefore imperative to predict how ongoing environmental changes will impact the functioning and stability of ecosystems (Hines et al. 2019).

A network that unites bottom-up and top-down processes in many ecosystems and is prone to strong alterations due to environmental change is the interaction network between parasitoids and their hosts (Tylianakis et al. 2006, Jeffs and Lewis 2013). Insect parasitoids attack and feed on and eventually kill their insect hosts (Godfray and Godfray 1994). Parasitoids are thought to be particularly sensitive to environmental changes, because species in higher trophic levels usually have smaller population sizes and their host fluctuations may cascade up to impact the parasitoids’ presence or absence. Therefore, insect host-parasitoid systems are ideal for studying the relationships between community-level changes and species interactions (Jeffs and Lewis 2013). Previous studies mainly focused on the influence of abiotic factors on host-parasitoid interactions, such as elevation and habitat structure (e.g. Valladares et al. 2012, Maunsell et al. 2015, Grass et al. 2018), or sometimes interaction structure (e.g. Cagnolo et al. 2011). However, the role of multiple components of plant diversity (i.e. taxonomic, functional and phylogenetic diversity) in modifying host-parasitoid interaction networks, as key biotic determinants of overall ecosystem structure, remains relatively little explored (but see Staab et al. 2016). Recent studies mainly focused on basic diversity associations between host and parasitoids (Ebeling et al. 2012, Schuldt et al. 2019, Guo et al. 2021). These studies have demonstrated both direct and indirect effects (i.e. one pathway and more pathways via other variables) of plant diversity on both hosts and parasitoids diversity, possibly via increased niche space and resource availability (Guo et al. 2021). Nevertheless, how these patterns propagate to their interaction networks is still unclear. Moreover, the effects of changing plant diversity are not always obvious when looking at plant species richness. The other diversity components (e.g. plant phylogenetic diversity) have been shown to better predict diversity-dependent bottom-up effects on host-parasitoid networks (e.g. Staab et al. 2016). It is especially important to take phylogenetic dependences (e.g. phylogenetic diversity or phylogenetic congruence) into account when it comes to the phylogenetic structure within and across trophic levels (Webb et al. 2002, Emerson and Gillespie 2008). This makes it vital to account for multiple dimensions of biodiversity (e.g. taxonomic, phylogenetic, functional) and relevant trophic interactions (Peralta et al. 2015, Volf et al. 2017, Wang et al. 2020). These components might jointly affect host-parasitoid networks in a system with high species diversity. Forests have garnered special attention lately, because they represent complex and large ecosystems susceptible to global change (De Frenne et al. 2021, Popkin 2021). Understanding how multiple dimensions of biodiversity modulate the effects of tree diversity loss on the structure and interaction strength of host-parasitoid networks clearly requires further study (Staab et al. 2016, Fornoff et al. 2019).

Here, we leverage standardized trap nests for solitary cavity-nesting bees, wasps, and their parasitoids in a large-scale subtropical forest biodiversity experiment to test how multiple dimensions of tree diversity and community composition influence host-parasitoid network structure. A multi-faceted approach is particularly important when considering that associations between trophic levels might be non-random and phylogenetically structured (Volf et al. 2018, Wang et al. 2020). We aimed to discern the primary components of the diversity and composition of tree communities that affect higher trophic level interactions via quantifying the strength and complexity of associations between hosts and parasitoid. We expected that (a) multiple tree community metrics, such as species, phylogenetic and functional diversity and species community composition can structure host and parasitoid community compositions, especially via phylogenetic processes (e.g. lineages of trophic levels diverge and evolve over time), as species interactions often show phylogenetic conservatism (e.g. Pellissier et al. 2013, Peralta et al. 2015). Further, we hypothesized that (b) host-parasitoid networks will be more complex and stable with increasing tree species richness due to potential links from higher richness of hosts and parasitoids promoted by more tree species, and (c) both, community and interaction network changes can also be related to abiotic factors, such as canopy cover, which might play a role in structuring hymenopteran communities (Haddad et al. 2011, Fornoff et al. 2021). By better understanding these dynamics, we can begin building a generalized framework for understanding host-parasitoid interactions in forest ecosystems.

Results

Overall, 34,398 brood cells were collected from 13,267 tubes across five years of sampling (2015, 2016, 2018, 2019, and 2020). Six families of hosts and seventeen families of parasitoids were identified. Among them, we found 56 host species (12 bees and 44 wasps, mean abundance and richness are 400 and 45, respectively, for each plot) and 50 parasitoid species (38 Hymenoptera and 12 Diptera, mean abundance and richness are 14 and 9, respectively, for each plot). The full species list and their abundances are given in Table S1. Overall, our sampling was adequate for analysis (especially for hosts), as confirmed by the sampling completness evaluation (Fig. S1).

Community composition of hosts and parasitoids

Host species composition was significantly related to the species and phylogenetic composition of the trees and parasitoid communities (NMDS axis scores; i.e. preserving the rank order of pairwise dissimilarities between samples), as well as to canopy cover, tree mean phylogenetic distance (MPD), elevation, and eastness (sine-transformed radian values of aspect) (Fig. 1a, Table 1, Table S2). Parasitoid species composition was significantly associated with host phylogenetic diversity, tree functional diversity, tree MPD, eastness, and elevation, and was significantly related to tree species composition, host species composition, and canopy cover (Fig. 1b, Table 1, Table S3). Host phylogenetic composition was affected by tree species composition, tree MPD, tree functional diversity, canopy cover, eastness, elevation and was especially affected by parasitoid species and phylogenetic composition (Fig. 1c, Table 1, Table S4). For parasitoid phylogenetic composition, significant relationships were found with tree species and phylogenetic composition, host species composition, tree functional diversity, canopy cover, elevation, and eastness (Fig. 1d, Table 1, Table S5). The PERMANOVA analysis also highlighted the important role of canopy cover for host and parasitoid community (Table S6-9). The Mantel test revealed a consistent pattern with the NMDS analysis, highlighting a pronounced relationship between the species composition of hosts and parasitoids (Table S10). However, the correlation between the phylogenetic composition of hosts and parasitoids was not significant.

Ordination plot of the non-metric multidimensional scaling (NMDS) analysis of (a) host species composition, (b) parasitoid species composition, (c) host phylogenetic composition, and (d) parasitoid phylogenetic composition across the study plots (filled circles) in the BEF-China experiment. Stress = 0.23, 0.23, 0.24 and 0.20, respectively. Arrows indicate significant (at p < 0.05) correlations of environmental variables with NMDS axis scores. Lengths of arrows are proportional to the strength of the correlations. Red crosses refer to the host or parasitoid species in each community. See Table S2-S5 in the Supplementary Material for abbreviations and statistical values.

An effect of host composition on the composition of the parasitoid communities was further indicated by a significant parafit test (p = 0.032) for testing the hypothesis of coevolution between a clade of hosts and a clade of parasites, suggesting nonrandom associations in the phylogenetic structure of parasitoid and host communities (Fig. 2, Fig. S4).

Dendrogram of phylogenetic congruence for the host species (below) and associated parasitoid species (above) recorded in the study. Each rectangle represents a different superfamily (for host species) or family (for parasitoid species). H1: Pompilidae H2: Apoidea H3: Vespidae; P1: Sarcophagidae P2: Phoridae P3: Bombyliidae P4: Trigonalyidae P5: Mutillidae P6: Megachilidae P7: Chrysididae P8: Ichneumonidae P9: Chalcidoidea. The trophic network of hosts and parasitoids was non-randomly structured (parafit test: P = 0.032). Host and parasitoid species names are given in Fig. S4.

Host-parasitoid network associations

The linear regression model results showed that host vulnerability, linkage density, and robustness of parasitoids were significantly negatively related to tree species richness, while remaining unaffected by other environmental covariables (Table 2, Fig. 3; except for elevation, which was marginally significantly related to robustness). Interaction evenness was significantly negatively associated with canopy cover, and interaction evenness was also negatively related to eastness (Fig. 4c; Table 2). Parasitoid generality was only marginally associated with canopy cover, and was not related to tree species richness or the other environmental variables. In the alternative models (tree species richness replaced by tree MPD), host vulnerability and linkage density were significantly positively related to tree MPD (Fig. 4a, 4b; Table S11), while robustness of parasitoids was negatively related to tree MPD (Fig. S5, Table S11). The results of other network metrics (parasitoid generality and interaction evenness) were consistent with those of the primary models. Tree mean nearest taxon distance (MNTD) was unrelated to any network indices.

Community-level relationships of network between tree species richness and (a) vulnerability, (b) linkage density, and (c) robustness of parasitoids. Values were adjusted for covariates of the final regression model. Regression lines (with 95% confidence bands) show significant (p < 0.05) relationships. Note that axes are on a log scale for tree species richness.

Community-level relationships of network between tree mean pairwise phylogenetic distance and (a) vulnerability and (b) linkage density and community-level relationships of network between canopy cover and (c) interaction evenness. Values were adjusted for covariates of the final regression model. Regression lines (with 95% confidence bands) show significant (p < 0.05) relationships.

The results of the null model analysis suggested that our metrics calculated by the observed network were significantly different from a random distribution (72, 71, and 77 out of 85 values for parasitoid generality, host vulnerability, and linkage density, respectively; all values for robustness, and interaction evenness), strongly demonstrating that interactions between species were not driven by random processes.

Discussion

Our study demonstrates that tree species richness and phylogenetic diversity play key roles in modulating interacting communities of hosts and parasitoids. These interactions are further structured by the phylogenetic associations between hosts and parasitoids. Moreover, canopy cover partly determined host-parasitoid network indices patterns, including host vulnerability, linkage density, and interaction evenness. This result supports a recent finding that the structure of host-parasitoid networks is also mediated by changes in microclimate, which is directly related to canopy cover, to some extent (Fornoff et al. 2021). These patterns were highly associated with multiple tree diversity metrics (tree species, phylogenetic and functional diversity), and compositional changes which are key to understand how host-parasitoid interactions may be impacted by biodiversity loss of lower trophic levels in food webs through trait- and phylogeny-based processes.

Community composition associations

Host species composition was influenced by several factors, including the species and phylogenetic composition of trees and parasitoids, tree diversity (species richness and MPD), and other biotic factors such as elevation and slope (Fig. 1a; Table 1). The effects of tree diversity and composition on host species composition agree with previous studies where solitary bee and wasp species composition were related to plant community structure (e.g. Loyola and Martins 2008). It seems likely that these results are based on bee linkages to pollen resources and predatory wasp linkages to the diverse of food sources, which may themselves be closely linked to resource heterogeneity increasing with tree species richness (Reitalu et al. 2019, Staab and Schuldt 2020).

We also found that tree MPD, FD, and species composition affect parasitoid species composition (Fig. 1b; Table 1), as many studies have found significant relationships between plant and parasitoid diversity in forests, including tree phylogenetic diversity (Staab et al. 2016), functional diversity (Rodriguez et al. 2019), and structural diversity (Schuldt et al. 2019). Similar to predators, parasitoids might also be more active or efficient with increasing tree community dissimilarity due to higher prey resources or lower intraguild parasitism caused by more diverse habitats (Finke and Denno 2002). On the other hand, our results also show that host species composition and parasitoid species composition relate to each other and their phylogenetic compositions, which are structured by tree communities to some extent. This pattern could propagate to the adjacent two higher trophic interactions through both top-down and bottom-up control.

Both host and parasitoid phylogenetic composition were related to tree species composition (Fig. 1c, d; Table 1). This pattern has important implications for cascading effects among trophic levels, in that producer communities could structure a higher trophic level community via an intermediate trophic level. However, while both host and parasitoid phylogenetic composition was related to tree MPD, only parasitoids responded to tree phylogenetic composition. This may be because there are many caterpillar-hunting wasps in our host communities, and the community composition of caterpillars were usually correlated with tree phylogenetic communities (Wang et al. 2019). Therefore, the prey organisms highly associated with tree phylogenetic composition (e.g. caterpillars) might indirectly determine predatory wasp (host) phylogenetic composition, as recently found for interactions between plants, caterpillars and spiders (Chen et al. 2023). This could be further tested by analyzing the food directly used by the wasps (caterpillars). For parasitoids, tree phylogenetic composition might drive the process of community assembly through trophic cascades (e.g. from plants to parasitoids via herbivores and host wasps) (Webb et al. 2002, Cavender-Bares et al. 2009). Additionally, parasitoid phylogenetic composition can be indirectly influenced by tree structural diversity (e.g. host availability in plots with higher heterogeneity; Schuldt et al. 2019), which can be determined by conserved traits across phylogenies (Webb et al. 2002). The phylogenetic associations between hosts and parasitoids exhibited a nonrandom structure (significant parafit correlation; Fig. 2) between the phylogenetic trees of the host and their parasitoids (see also Peralta et al. 2015). Such pattern could be further confirmed by the significant association between host phylogenetic composition and parasitoid phylogenetic composition (Fig. 1c), which suggested that their interactions are phylogenetically structured to some extent. However, this significant pattern was observed only in the NMDS analysis and not in the Mantel test, suggesting that the non-random interactions between hosts and parasitoids could not be simply predicted by their community similarity and associations between the phylogenetic composition of hosts and parasitoids are more complex and require further investigation in the future.

Moreover, the species and phylogenetic composition of hosts and parasitoids was also related to abiotic factors, especially to canopy cover, which has been considered especially important to microclimate (Sobek et al. 2009, Fornoff et al. 2021). In future studies, it will be useful to incorporate other, more direct metrics of microclimate, such as local temperature and humidity, to determine the proximal drivers of these microclimatic effects (Ma et al. 2010, Fornoff et al. 2021).

Community-level host-parasitoid networks

Tree species richness did not significantly influence the diversity of hosts targeted by parasitoids (parasitoid generality), but caused a significant increase in the diversity of parasitoids per host species (host vulnerability) (Fig. 3a; Table 2). This is likely because niche differentiation often influences network specialization via potential higher resource diversity in plots with higher tree diversity (Lopez-Carretero et al. 2014). Such positive relationship between host vulnerability and tree species richness suggested that host-parasitoid interactions could be driven through bottom-up effects via benefit from tree diversity. For example, parasitoid species increases more than host diversity with increasing tree species richness (Guo et al. 2021), resulting in an increase of host vulnerability at community level. According to the enemies hypothesis (Root 1973), which posits a positive effects of plant richness on natural enemies, the higher trophic levels in our study (e.g. predators and parasitoids) would benefit from tree diversity and regulate herbivores thereby (Staab and Schuldt 2020). Indeed, previous studies at the same site found that bee parasitoid richness and abundance were positively related to tree species richness, but not their bee hosts (Fornoff et al. 2021, Guo et al. 2021). Because our dataset considered all hosts and reflects an overall pattern of host-parasitoid interactions, the effects of tree species richness on parasitoid generality might be more complex and difficult to predict, as we found that neither tree species richness nor tree MPD were related to parasitoid generality.

Linkage density was positively related to tree species richness (Fig. 3b; Table 2), supporting the food web theory, which predicts in our case that network complexity (linkage density) depends on the number of plants (Blüthgen and Staab 2024). Although trees were not directly included as a trophic level in our networks, potential network complexity increased with tree species richness, likely enabling higher network stability/resistance (Ebeling et al. 2011, Staab et al. 2015). For example, a network might be more sensitive to extinctions because of key species loss due to lower linkage density and lower redundancy (Blüthgen and Staab 2024). However, parasitoid robustness was unexpectedly negatively related to tree species richness (Fig. 3c; Table 2). It was expected that higher trophic levels would be more robust, less influenced by perturbations from lower trophic levels, when plant diversity is higher, as more potential interactions at lower trophic levels should theoretically increase redundancy and resilience of connected higher levels (Blüthgen and Klein 2011, Fornoff et al. 2019). Dilution effects may explain this, as plots with higher richness held fewer individuals of a given tree species. If there are strong prey item (caterpillars, grasshoppers, etc.) preferences for one species, there may be fewer individuals per area or they may be more densely aggregated and less likely to be encountered by parasitoids. This increased stochasticity in parasitoid wasps could benefit hosts by reducing parasitism pressure overall, weakening top-down controls.

Similar to tree species richness, tree MPD was also positively correlated with host vulnerability and linkage density (Fig. 4a, b), meaning that the mean number of parasitoids per host species and number of links within the host-parasitoid system can also be promoted by tree MPD. This is in agreement with several recent studies showing that plant phylogenetic diversity not only affects herbivores but also higher tropic levels (Pellissier et al. 2013, Staab et al. 2020, Wang et al. 2020). Our results suggest that the specialization and complexity of higher trophic levels can also be affected by plant phylogenetic diversity. This pattern can be traced to the effects of habitat heterogeneity caused by tree species richness and MPD on higher trophic levels via bottom-up control. The effects of tree MPD were consistent with effects of tree species richness on robustness of parasitoids to host loss. This result suggests that higher trophic levels are sensitive to changes in both plant phylogenetic relatedness and general species dissimilarity via trophic interactions, even the hosts are not all directly interacting with plants, bees excluded. Therefore, it may be that stronger linkages would be found when exclusively exploring such plant-herbivore-parasitoid systems.

Interaction evenness was significantly negatively related to canopy cover(Fig. 4c), further reinforcing the important role of canopy-cover modulated microclimate (likely temperature and humidity) for trophic interactions (Sobek et al. 2009, Fornoff et al. 2021). Our results agree with a previous study on ants, where plant-insect interactions were more even with more open canopies (Dáttilo and Dyer 2014). In our case, canopy cover might change hymenopteran species evenness and then further influence interaction evenness. Certain host species tended to nest in plots with higher canopy cover, which might decrease the interaction evenness by favoring parasitoids of fewer, more dominant hosts. This pattern would become more significant when more host and parasitoid species are in a plot, given the positive relationship between higher trophic level diversity and canopy cover.

Future prospects

Overall, our study enables new insights into the dynamics of host-parasitoid interactions under varying environmental conditions, an important step toward building a synthetic model for such biodiversity. A key finding was that although parasitoids and hosts respond to tree species richness, top-down control seems predominant in parasite-host interaction, though whether this holds for others antagonistic interactions requires further investigation. Different trophic levels and functional groups of species responded differently to experimental changes in plant communities (Fornoff et al. 2021, Guo et al. 2021). This highlights the complexities of building multi-trophic networks and calls for more studies across habitat types and taxa, to test the generality of our findings. Future studies should also consider the role of host/parasitoid functional traits, because they might play a critical role in modifying network structures and ecosystem functioning.

Materials and methods

Study sites design

This study was conducted in the BEF-China biodiversity experiment, which is the largest tree diversity experiment worldwide. The experiment is located in a subtropical forest near Xingangshan, Jiangxi province, south-east China (29°08′–29°11′N, 117°90′–117° 93′E). The mean annual temperature is 16.7°C and mean annual precipitation 1821 mm (Yang et al. 2013). The experiment includes two study sites (Site A and Site B), 4 km apart from each other, that were established in 2009 (Site A) and 2010 (Site B) respectively. A total of 566 plots (25.8×25.8 m) were designed on the two sites, and per plot 400 trees were initially planted in 20 rows and 20 columns with a planting distance of 1.29 m. A tree species richness gradient (1, 2, 4, 8, 16 and 24 species) was established at each site, based on a species pool of 40 local, broadleaved tree species (Bruelheide et al. 2014). The tree species pools of the two plots are nonoverlapping (16 species for each site). The composition of tree species within the study plots is based on a “broken-stick” design (see Bruelheide et al. 2014).

For our study, at both sites (site A and site B) eight plots of each tree species richness level (1, 2, 4, 8) were randomly selected, as well as six and two plots of 16 and 24 mixtures. In addition, at site B eight additional monocultures were sampled (Fornoff et al. 2021), resulting in 48 plots in Site B (including 16 monocultures, eight plots for each 2, 4, 8 mixtures and six and two plots of 16 and 24 mixtures. In total, 88 study plots were used (40 plots on Site A and 48 plots on Site B, see Fig. S2).

Sampling

We collected trap nests monthly to sample solitary bees and wasps (Staab et al. 2018) in the 88 plots from September to November in 2015 and April to November in 2016, 2018, 2019 and 2020. For each plot, we installed two poles with trap nests (11 m apart from each other and 9 m away from the nearest adjacent plots) along a SW–NE diagonal (following the design of Ebeling et al. 2012). Each pole stood 1.5 m above ground, and each trap nest consisted of two PVC tubes (length: 22 cmL×Ldiameter: 12.5 cm) filled with 75L±L9 (SD) reed internodes of 20 cm length and diameters varying between 0.1 and 2.0 cm (Staab et al. 2014, Fornoff et al. 2021). Every month, we sampled the reeds with nesting hymenopterans and replaced them with internodes of the same diameter. All the samples were reared in glass test tubes under ambient conditions until specimens hatched. We identified hatched hosts and parasitoids to species or morphospecies (Supplementary Table S1) based on reference specimens (vouchered at the Institute of Zoology, CAS, Beijing). We were interested in the general patterns of host-parasitoid interactions at the community level, so for the analysis we did not distinguish between the two life-history strategies of parasitoids (true parasitoids and kleptoparasitoids, including hymenopteran and dipteran parasitoids) because they both have the same ecological result, death of host brood cells. We evaluated our sampling completeness with r package “iNEXT” (Hsieh et al. 2016).

DNA extraction and amplification

All specimens were sequenced for a region of the mitochondrial cytochrome c oxidase subunit I (COI) gene (Hebert et al. 2003). We extracted whole-genomic DNA of hosts and parasitoids using DNeasy Blood & Tissue Kits (QIAGEN GmbH, Hilden, Germany), following the manufacturer’s protocols. COI sequences of samples were amplified using universal primer pairs, LCO1490 (GGTCAACAAATCATAAAGATATTGG) as the forward primer and HCO2198 (TAAACTTCAGGGTGACCAAAAAATCA) or HCOout (CCAGGTAAAATTAAAATATAAACTTC) as the reverse primer. We carried out polymerase chain reactions (PCR) in 96-well plates with 30 μl reactions containing 10 μl ddH2O, 15 μl Premix PrimeSTAR HS (TaKaRa), 1 ul of each primer at 10 μM, and 3 μl template genomic DNA using a thermo cycling profile. The PCR procedure as follows: 94℃ for 1 min; 94℃ for 1 min, 45℃ for 1.5 min and 72℃ for 1.5 min, cycle for 5 times; 94℃ for 2 min, 58℃ for 1.5 min and 72℃ for 1 min, cycle for 36 times; 72℃ for 5 min. We performed all PCRs on an Eppendorf Mastercycler gradient, which were then visualized on a 1% agarose gel. Samples with clean single bands were sequenced after PCR purification using BigDye v3.1 on an ABI 3730xl DNA Analyser (Applied Biosystems).

Sequence alignment and phylogenetic analysis

We applied MAFFT (Misawa, Katoh, Kuma, & Miyata, 2002) to align all sequences, then translated the nucleotides into amino acids via MEGA v7.0 (Kumar, Stecher, & Tamura, 2016) to check for the presence of stop codons with manual adjustments. Host and parasitoid sequences were then aligned against the references using a Perl-based DNA barcode aligner (Chesters, 2019).

We employed two strategies to improve the phylogenetic structure of a DNA barcode phylogeny, which demonstrably improve resulting phylogeny-based diversity indices (Macías-Hernández et al. 2020). These include the integration of 1) molecular sequences of the plot data and 2) phylogenetic relationships from other molecular datasets. Integration was achieved following Wang et al. (2020) and Chesters (2020): reference DNA barcodes of Hymenoptera and Diptera were downloaded from the BOLD API (www.boldsystems.org/index.php/API_Public), which were variously processed (e.g. to retain only fully taxonomically labelled barcodes, to remove low quality or mislabeled entries, and to dereplicate to a single exemplar per species), and then aligned (Chesters 2019). A single outgroup was included for which we selected the most appropriate insect order sister to Diptera and Hymenoptera (Misof et al. 2014), a representative of the order Psocoptera (Psocidae, Psocus leidyi). We then constructed a phylogeny of the references and subjects, with references constrained according to the method described earlier (Chesters 2020). A number of backbone topologies were integrated for setting hard and soft constraints, including a transcriptomics-derived topology (Chesters 2020), a mitogenome tree of insects (Chesters 2017), Diptera-specific trees (Wiegmann et al. 2011, Cranston et al. 2012, Ament 2017) and Hymenoptera-specific trees (Peters et al. 2011, Branstetter et al. 2017, Cardinal 2018). The constrained inference was conducted with RaxML version 8 (Stamatakis 2014) under the standard GTRGAMMA DNA model with 24 rate categories. According to the backbone trees used, most taxa present were monophyletic with a notable exception of Crabronidae, for which there is emerging phylogenomic evidence of its polyphyly (Sann et al. 2018).

Tree phylogenetic diversity, functional diversity, and environmental covariates

The phylogenetic diversity of the tree communities was quantified by calculating wood volume-weighted phylogenetic Mean Pairwise Distance (MPD) (Tucker et al. 2017). Tree wood volume was estimated from data on basal area and tree height (Bongers et al. 2021) measured in the center of each plot. Moreover, to represent variations towards the tips of the phylogeny beyond MPD, we additionally calculated Mean Nearest Taxon Distance (MNTD), which is a measure that quantifies the distance between each species and its nearest neighbor on the phylogenetic tree (Webb 2000). Phylogenetic metrics of trees (tree MPD and MNTD) were calculated based on a maximum likelihood phylogenetic tree available for the tree species in our study area (Purschke et al. 2017). Considering that predatory wasps mainly feed on herbivorous caterpillars, we calculated tree functional diversity to test the indirect effects on hymenopteran communities and relevant network indices. Specifically, seven leaf traits were used for calculation of tree functional diversity, which was calculated as the mean pairwise distance in trait values among tree species, weighted by tree wood volume, and expressed as Rao’s Q (Ricotta and Moretti 2011), including specific leaf area, leaf toughness, leaf dry matter content, leaf carbon content, ratio of leaf carbon to nitrogen, leaf magnesium content, and leaf calcium content. These functional traits were commonly related to higher trophic levels in our study area, such as herbivores and predators (Wang et al. 2020, Chen et al. 2023), which are the main food resources of our predatory wasps. All of the traits were measured on pooled samples of sun-exposed leaves of a minimum of five tree individuals per species following standard protocols (Pérez-Harguindeguy et al. 2003).

As our analyses mainly compare community patterns among study plots, we additionally considered potential effects of environmental variation by using plot means of slope, elevation, “eastness” (sine-transformed radian values of aspect), and “northness” (cosine-transformed radian values of aspect) as environmental covariates that characterize the heterogeneity of the study plots. We also accounted for the potential effects of canopy cover at plot level for host-parasitoid interactions, as it can structure hymenopteran communities (Perlík et al. 2023). Canopy cover was calculated as in Fornoff et al. (2021) based on hemispherical photographs.

Statistical analysis

All analyses were conducted in R 4.1.2 with the packages ape, vegan, picante, bipartite, and caper (http://www.R-project.org). Prior to analysis, samples from the five years (2015, 2016, 2018, 2019, and 2020) were pooled at the plot level to discern overall and generalizable effects permeating this system. We excluded three plots with no living trees because of high mortality, resulting in 85 plots in the final analysis.

Composition of trees, hosts and parasitoids

The species and phylogenetic composition of trees, hosts, and parasitoids were quantified with nonmetric multidimensional scaling (NMDS) analysis based on Morisita-Horn distances. The minimum number of required dimensions in the NMDS based on the reduction in stress value was determined in the analysis (k = 2 in our case). We centered the results to acquire maximum variance on the first dimension, and used the principal components rotation in the analysis. The phylogenetic composition was calculated by mean pairwise distance among the host or parasitoid communities per plot with the R package “picante” applying the ‘mpd’ function. To test the influence of study plot heterogeneity on these relationships, we fitted their standardized values (see vectors in Table S2) to the ordination on the basis of a regression with the NMDS axis scores (Quinn and Keough 2002). NMDS was widely used to summarize the variation in species composition across plots. The two axes extracted from the NMDS represent gradients in community composition, where each axis reflects a subset of the compositional variation. These axes should not be interpreted in isolation, as the overall species composition is co-determined by their combined variation. For clarity, results were interpreted based on the relationships of variables with the compositional gradients captured by both axes together. For the analysis, we considered tree species richness, tree functional and phylogenetic diversity, canopy cover, and environmental covariates (elevation, eastness, northness, and slope) as plot characteristics. We assessed the significance of correlations with permutation tests (permutation: n= 999). To strengthen the robustness of our findings based on NMDS, we further validated the composition results using Mantel test and PERMANOVA (with ‘adonis2’) for correlation between communities and relationships between communities and environmental variables.

Phylogenetic match of hosts and parasitoids

In addition, we used a parafit test (9,999 permutations) with the R package “ape” to test whether the associations were non-random between hosts and parasitoids. This is widely used to assess host-parasite co-phylogeny by analyzing the congruence between host and parasite phylogenies using a distance-based matrix approach. The species that were not attacked by parasitoids or failed to generate sequences were excluded from the analyses. For species abundance and composition, see Table S1.

Host-parasitoid interactions

We constructed quantitative host-parasitoid networks at community level with the R package “bipartite” for each plot of the two sites. Bees and wasps were considered together as hosts because there were too few abundant bee species to analyze separate interaction networks for bees and wasps. We calculated five indices to quantitatively characterize the structure of the interaction networks (Blüthgen and Staab 2024): weighted parasitoid generality (effective number of host species per parasitoid species), weighted host vulnerability (effective number of parasitoid species attacking a host species), robustness (degree of network stability), linkage density (degree of network specialization), and interaction evenness (degree of network evenness). Parasitoid generality was defined as the weighted mean number of host species per parasitoid species, with A being the number of interactions of parasitoid species j, m the total number of interactions of all species, and Hj the Shannon diversity of interactions of species j. Host vulnerability was the weighted mean number of parasitoid species per host species, Vulnerability = (Bersier et al. 2002). i=l m Robustness was defined as the area under the extinction curve, reflecting the degree of decreases of one trophic level with the random elimination species of the other trophic levels, here using the robustness index for higher trophic levels (i.e. parasitoids). The robustness calculation was also weighted based on interaction abundance. For linkage density, Lq = we used the realized proportion of possible links between the two trophic levels as the mean number of interactions per species across the entire network (Tylianakis et al. 2007). Interaction evenness was defined as ES = - ∑ i∑jpij lnpij/ln/J, which is used to describe Shannon’s evenness of network interactions (Dormann et al. 2009). To check whether all network indices significantly differ from chance across all study plots, we used Patefield null models (Dormann et al. 2009) to compare observed indices with simulated values (10,000 times).

Linear models

To test the effects of tree species richness, tree phylogenetic, and functional diversity, as well as canopy cover and the other environmental covariates (including slope, elevation, eastness, and northness) on the six network indices, we used linear mixed-effect models. For our analyses, the study sites were considered as random effects, and the other variables were treated as fixed effects (see above). Given the strong correlation between tree species richness and tree MPD (Pearson’s r = 0.74, p<0.01), we excluded tree MPD in the models where tree species richness was a predictor. To evaluate the potential effects caused by tree MPD, we also ran alternative models where tree species richness was replaced with tree MPD. We simplified all models by gradually removing non-significant factors to obtain the most parsimonious model with the lowest AICc. To ensure that the analyses were not strongly affected by multicollinearity, the correlations among all predictors were tested (Fig. S3), and variance inflation factors (VIF) of our statistical models were checked.